---
title: "Report of Parameters used with V-SVA"
output: html_document
date: "`r Sys.time()`"
fig_width: 12
fig_height: 12 
params:
  dt: NA
---

##Expression Data and Sample Metadata Provided
```{r}
# dimensions of expression matrix
if (!is.null(params$dt$exp_norm)) {
  dim(params$dt$exp_norm)
}
# dimensions of metadata, if not provided only Genes_Detected and Log_Total_Counts determined
if (!is.null(params$dt$meta_df)) {
  dim(params$dt$meta_df)
  head(params$dt$meta_df)  
}

```

##Quality Control of Expression Data: Number of Features Detected in each Sample
```{r, eval=FALSE}
# binarize data to determine which features are detected in each sample
bin_data <- params$dt$exp_norm
bin_data[bin_data < 1] <- 0
bin_data[bin_data >= 1] <- 1
num.exp <- apply(bin_data,2,sum)
params$dt$detect_num <- num.exp
```

```{r}
if (!is.null(params$dt$detect_num)) {
  summ <- summary(params$dt$detect_num)
  # histogram of number of features detected in each sample
  hist(params$dt$detect_num, col = "dodgerblue", main="", 
     ylab = "Samples (n)", xlab = "Number of features detected in each sample")
  legend("topright", legend = paste(names(summ), round(summ, digits = 2), sep = " "), title = "Summary of features detected")
}
```

##Feature Pre-processing Options (normalization, sample and feature filtering)
#### Remove samples based on number of detected features
```{r}
# were cells filtered based on genes detected?
if (params$dt$cell_filter_choice) {
 print(paste("Removing cells that have less than ", params$dt$cellfilt_number, " features detected,", sep = "")) 
} else {
  print("Cells were not filtered based on features detected")
}
```

```{r, eval=FALSE}
# code to filter cells based on features detected
num.sel <- params$dt$detect_num[params$dt$detect_num >= params$dt$cellfilt_number]
# subset data
params$dt$exp_norm <- params$dt$exp_norm[, names(num.sel)]
params$dt$meta_df <- params$dt$meta_df[names(num.sel), ]

```

#### Down-sample the number of samples included in the analysis (to increase computational efficiency)
```{r}
# were cells down-sampled to a certain number to speed up computation?
if (params$dt$cell_downsample_choice) {
 print(paste("Cells were down-sampled to ", params$dt$cell_downsample_number, " cells", sep = "")) 
} else {
  print("Cells were not down-sampled")
}
```

```{r, eval=FALSE}
# code to down-sample cells
set.seed(1)
dw_samp <- base::sample(x = 1:ncol(params$dt$exp_norm), size = params$dt$cell_downsample_number)
# subset data
params$dt$exp_norm <- params$dt$exp_norm[, dw_samp]
params$dt$meta_df <- params$dt$meta_df[dw_samp, ]
```

#### Remove features not detected in samples
```{r}
# were features filtered?
if (params$dt$gene_filter_choice) {
 print(paste("Features were filtered using ", 
             params$dt$gene_filter_method,
             ".", nrow(params$dt$exp_norm), " Features with ", params$dt$gene_count_num, 
             " or more counts in at least ", params$dt$gene_cell_num,
             " cells were retained" , sep = "")) 
} else {
  print("Features were not filtered based on detection rate in samples")
}
    
```

#### Choose normalization method for expression data
```{r, eval=FALSE}
# code for gene filtering methods: all code is displayed regardless of method chosen
filter <- apply(params$dt$exp_norm, 1, function(x) length(x[x>isolate(input$Count_num)])>=isolate(input$Cell_num))
params$dt$exp_norm <- params$dt$exp_norm[filter,]
# normalize the data
# using CPM method
if (isolate(input$norm_method) == "CPM") {
  params$dt$exp_norm <- edgeR::cpm(params$dt$exp_norm)
  # using quantile normalization method
} else if (isolate(input$norm_method) == "Quantile") {
  params$dt$exp_norm <- normalize.quantiles(params$dt$exp_norm)
  # scran method
} else if (isolate(input$norm_method) == "scran") {
  sce <- SingleCellExperiment(list(counts=params$dt$exp_norm))
  sce <- computeSumFactors(sce)
  sce <- normalize(sce)
  params$dt$exp_norm <- exprs(sce)
  # no normalization
} else if (isolate(input$norm_method) == "None") {
  params$dt$exp_norm <- params$dt$exp_norm
}
```

##Surrogate Variable Analysis (SVA)
```{r}
# which method was used
if (!is.null(params$dt$sva_method_use)) {
  print(paste("The surrogate variable analysis method chosen was: ", params$dt$sva_method_use, sep = ""))  
}

# which known factors were adjusted for
if (!is.null(params$dt$known_factors_use)) {
  print(paste("The following known factor(s) were adjusted for: ", params$dt$known_factors_use, sep = ""))
}

# the number of SV's 
if (!is.null(params$dt$iasva.res)) {
  print(paste("The number of SV's/Factors identified: ", ncol(params$dt$iasva.res$sv), sep = ""))
}
```

```{r, eval=FALSE}
# code for SVA methods
# create model matrix with known factors to adjust for
id_mod <- which(colnames(params$dt$meta_df) %in% params$dt$known_factors_use)
if (length(id_mod) > 1) {
  formdf1 <- as.formula(paste("~", colnames(params$dt$meta_df)[id_mod][1], "+", paste(colnames(params$dt$meta_df)[id_mod[2:length(id_mod)]],collapse="+"), sep = ""))
  mod <- model.matrix(formdf1, data = params$dt$meta_df)
} else {
  varf1 <- as.factor(params$dt$meta_df[, id_mod])
  mod <- model.matrix(~varf1, data = params$dt$meta_df)
}
# create summarized experiment for expression matrix to later use for marker gene identification
summ_exp <- SummarizedExperiment(assays = as.matrix(params$dt$exp_norm))
params$dt$summ_exp <- summ_exp
      
# if user chose IA-SVA, then perform following
if (isolate(params$dt$sva_method_use == "IA-SVA")) {
# depending on which ia-sva parameters were chosen, evaluate
  if (isolate(input$iasva_param == "Percentage Threshold")) {
    params$dt$iasva.res <- fast_iasva(summ_exp, mod[,-1, drop = F], verbose=FALSE,
                                      pct.cutoff = isolate(input$pct_cutt), num.sv = NULL)
  } else if (isolate(input$iasva_param == "Number of SVs")) {
    params$dt$iasva.res <- fast_iasva(summ_exp, mod[,-1, drop = F], verbose=FALSE,
                                      pct.cutoff = isolate(input$pct_cutt), num.sv = isolate(input$num_of_svs))
  } 
  # else if choose SVA method
  } else if (isolate(params$dt$sva_method_use == "SVA")) {
    # perform sva analysis with specified svs
    sva.res <- svaseq(params$dt$exp_norm, mod = mod, mod0 = mod[,1], n.sv = isolate(input$sva_num))
    colnames(sva.res$sv) <- paste("SV", 1:ncol(sva.res$sv), sep = "")
    params$dt$iasva.res <- sva.res
        
  # else if choose zinb-wave method
  } else if (isolate(params$dt$sva_method_use == "ZINB-WaVE")) {
    # perform analysis with specified latent factors
    zinb.matrix <- params$dt$exp_norm
    # coerce to integer
    mode(zinb.matrix) <- "integer"
    zinb.res <- zinbFit(Y = zinb.matrix, X = mod[,-1, drop = F], K = isolate(input$zinb_num))
    # extract factors
    zinb.fac <- getW(zinb.res)
    colnames(zinb.fac) <- paste("SV", 1:ncol(zinb.fac), sep = "")
    params$dt$iasva.res <- list(sv = zinb.fac)
  }
      
```

## Correlation Plot of SV's and Sample Metadata
```{r}
# change factors to numeric for correlation
if (!is.null(params$dt$meta_df) & !is.null(params$dt$iasva.res)) {
  meta_sel <- params$dt$meta_df
for (jcol in 1:ncol(meta_sel)) {
  meta_sel[,jcol] <- as.numeric(as.factor(meta_sel[,jcol]))
}
iasva_vars <- cbind(params$dt$iasva.res$sv, meta_sel)
# need to append column names to matrix
colnames(iasva_vars) <- c(paste("SV", 1:ncol(params$dt$iasva.res$sv), sep = ""),
                          colnames(params$dt$meta_df))
col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(abs(cor(iasva_vars)), type = "upper", method = "color",
         col = col(200), number.cex = 1,
         addCoef.col = "black",
         tl.col = "black", tl.srt = 90, diag = FALSE)
}

```

## Paired SV Plots
```{r}
if (!is.null(params$dt$iasva.res)) {
  iasva.sv <- as.data.frame(params$dt$iasva.res$sv)
rownames(iasva.sv) <- colnames(params$dt$exp_norm)
pairs(iasva.sv, main="", pch=20, cex=0.5, lower.panel = NULL)
}

```

## Interactive Paired SV Plots
```{r}
if (!is.null(params$dt$iasva.res)) {
  iasva.sv <- as.data.frame(params$dt$iasva.res$sv)
  rownames(iasva.sv) <- colnames(params$dt$exp_norm)
  plot_ly(iasva.sv, x = ~SV1, y = ~SV2, type = "scatter",
          mode = "markers", text = paste("Cell ID: ", rownames(iasva.sv), sep = ""),
          marker = list(
            opacity = 0.5
            )
  )
}
```

## Identifying Marker Features associated with SV's
```{r, eval=FALSE}
# identify which svs were chosen for marker analysis
id_sv_mark <- which(colnames(params$dt$iasva.res$sv) %in% isolate(input$SV_marks))
marker_genes <- iasva::find_markers(Y = params$dt$summ_exp,
                                    iasva.sv = as.matrix(params$dt$iasva.res$sv[, id_sv_mark, drop=FALSE]),
                                    rsq.cutoff = isolate(input$rsqcutoff), method = isolate(input$mark_sig), sig.cutoff = isolate(input$mark_cutoff))
params$dt$markers <- marker_genes
```

## Heatmap of Marker Features Determined from SVA Analysis
```{r}
if (!is.null(params$dt$markers)) {
  all_marks <- params$dt$markers$All_Unique_Markers
log_mat <- log(as.matrix(params$dt$exp_norm[all_marks,])+1)
# remove any NA's from matrix
log_mat <- log_mat[complete.cases(log_mat),]
pheatmap(log_mat, show_colnames = FALSE,
         show_rownames = TRUE,
         clustering_method = "ward.D2")
}
```

## Dimension Reduction
```{r, eval=FALSE}
set.seed(1)
# transpose matrix
trans_orig <- t(log(params$dt$exp_norm+1))
# remove any zeros
params$dt$pre_dim_orig <- trans_orig[, apply(trans_orig, 2, var, na.rm = TRUE) != 0]
# Principal component analysis (PCA) for all genes
dim_orig <- prcomp(x = params$dt$pre_dim_orig, center = TRUE, scale. = TRUE)
dim_orig_mat <- dim_orig$x
rownames(dim_orig_mat) <- colnames(params$dt$exp_norm)
params$dt$dim_orig <- as.data.frame(dim_orig_mat)

### Code for T-SNE ####
# T-SNE analysis (PCA) for all genes
# not comupted
# dim_orig <- Rtsne(X = params$dt$pre_dim_orig, dims = 3)
# dim_orig_mat <- dim_orig$Y
        
#### Code for MDS ###
# dim_orig <- cmdscale(d = dist(params$dt$pre_dim_orig), k = 3)
# dim_orig_mat <- dim_orig
# rownames(dim_orig_mat) <- colnames(params$dt$exp_norm)
# colnames(dim_orig_mat) <- c("MDS1", "MDS2", "MDS3")

# Principal component analysis for SV-selected genes
# transpose matrix
trans_mark <- t(params$dt$exp_norm[params$dt$markers_formatted[,1],])
# remove any zeros
params$dt$pre_dim_mark <- trans_mark[, apply(trans_mark, 2, var, na.rm = TRUE) != 0]
dim_mark <- prcomp(x = params$dt$pre_dim_mark, center = TRUE, scale. = TRUE)
dim_mark_mat <- dim_mark$x
rownames(dim_mark_mat) <- colnames(params$dt$exp_norm)
params$dt$dim_mark <- as.data.frame(dim_mark_mat)

```

### Dimension Reduction Visualization (using all features)
```{r}
if (!is.null(params$dt$dim_method)) {
  # print your dimension reduction method
  print(paste("Dimension reduction method chosen: ", params$dt$dim_method, sep = ""))

  # if chosen PCA 
  if (params$dt$dim_method == "PCA") {
    # 3D interactive dimension reduction plot using all features
    plot_ly(params$dt$dim_orig, x = ~PC1, y = ~PC2, z = ~PC3, type = "scatter3d",
          mode = "markers", text = paste("Cell ID: ", rownames(params$dt$dim_orig), sep = ""),
          marker = list(
            opacity = 0.5
          )) %>% layout(title = paste("All Genes (n = ", nrow(params$dt$exp_norm), ")", sep = ""))
  } else if (params$dt$dim_method == "t-SNE") {
    plot_ly(params$dt$dim_orig, x = ~tSNE1, y = ~tSNE2, z = ~tSNE3, type = "scatter3d",
            mode = "markers", text = paste("Cell ID: ", rownames(params$dt$dim_orig), sep = ""),
            marker = list(
              opacity = 0.5
            )) %>% layout(title = paste("All Genes (n = ", nrow(params$dt$exp_norm), ")", sep = ""))
  } else if (params$dt$dim_method == "Classical Metric MDS") {
    plot_ly(params$dt$dim_orig, x = ~MDS1, y = ~MDS2, z = ~MDS3, type = "scatter3d",
            mode = "markers", text = paste("Cell ID: ", rownames(params$dt$dim_orig), sep = ""),
            marker = list(
              opacity = 0.5
            )) %>% layout(title = paste("All Genes (n = ", nrow(params$dt$exp_norm), ")", sep = ""))
  }
}
```

### Dimension Reduction Visualization (using SV-associated features)
```{r}
if (!is.null(params$dt$dim_method)) {
  # 3D interactive dimension reduction plot using SV selected features
  if (params$dt$dim_method == "PCA") {
    plot_ly(params$dt$dim_mark, x = ~PC1, y = ~PC2, z = ~PC3, type = "scatter3d",
          mode = "markers", text = paste("Cell ID: ", rownames(params$dt$dim_mark), sep = ""),
          marker = list(
            opacity = 0.5
          )) %>% layout(title = paste(params$dt$sva_method_use, " Genes (n = ", nrow(params$dt$exp_norm[params$dt$markers_formatted[,1],]),
                                      ";",params$dt$chosen_svs, ")", sep = ""))
  } else if (params$dt$dim_method == "t-SNE") {
    plot_ly(params$dt$dim_mark, x = ~tSNE1, y = ~tSNE2, z = ~tSNE3, type = "scatter3d",
            mode = "markers", text = paste("Cell ID: ", rownames(params$dt$dim_mark), sep = ""),
            marker = list(
              opacity = 0.5
            )) %>% layout(title = paste(params$dt$sva_method_use, " Genes (n = ", nrow(params$dt$exp_norm[params$dt$markers_formatted[,1],]), "; ",
                                        params$dt$chosen_svs, ")", sep = ""))
  } else if (params$dt$dim_method == "Classical Metric MDS") {
    plot_ly(params$dt$dim_mark, x = ~MDS1, y = ~MDS2, z = ~MDS3, type = "scatter3d",
            mode = "markers", text = paste("Cell ID: ", rownames(params$dt$dim_mark), sep = ""),
            marker = list(
              opacity = 0.5
              )) %>% layout(title = paste(params$dt$sva_method_use, " Genes (n = ", nrow(params$dt$exp_norm[params$dt$markers_formatted[,1],]), "; ",
                                          params$dt$chosen_svs, ")", sep = ""))  
  }
}

```

## Gene Enrichment Analysis
```{r, eval=FALSE}
# convert gene symbols to Entrez ID (for example for human data)
gene.df <- bitr(gene, fromType = "SYMBOL",
                        toType = c("ENSEMBL", "ENTREZID"),
                        OrgDb = org.Hs.eg.db)
params$dt$species <- org.Hs.eg.db
params$dt$gene.df <- gene.df

# analysis for GO biological process terms (as an example)
ego <- enrichGO(gene = params$dt$gene.df$ENTREZID,
                OrgDb = params$dt$species,
                keyType = "ENTREZID",
                ont = "BP",
                pvalueCutoff = isolate(input$pvalue_cutoff), pAdjustMethod = isolate(input$pvalue_correct),
                qvalueCutoff = isolate(input$path_cutoff),
                minGSSize = 5,
                readable = TRUE)
params$dt$enrich_res <- ego
params$dt$category_number <- input$path_viz_num
params$dt$pathway_name <- input$Path_Type
```

```{r}
# visualize pathway results
if (!is.null(params$dt$enrich_res)) {
  print(paste("Type of gene enrichment analysis performed: ", params$dt$pathway_name, sep = ""))
  print(paste("Show this many categories: ", params$dt$category_number, sep = ""))

  dp <- clusterProfiler::dotplot(object = params$dt$enrich_res, showCategory = params$dt$category_number) + ggtitle(params$dt$pathway_name)
  plot(dp)
}
```


## Session Information
```{r}
sessionInfo()
```

